---
title: "HW-1"
format: pdf
---

```{r}
library(palmerpenguins) 
```
# line 7: loading in the library

```{r}
dim(penguins) 
str(penguins) # variable types present in the dataset.
summary(penguins) 
```
# line 12: number of rows and columns in the dataset. there are 344 rows and 8 columns.
# line 13: shows more information about all the variables and their corresponding set of values.

```{r}
# 1.1: Extract and analyze bill lengths
bill_lengths <- penguins[["bill_length_mm"]]

mean(bill_lengths, na.rm = TRUE) 
median(bill_lengths, na.rm = TRUE)
sd(bill_lengths, na.rm = TRUE)
min(bill_lengths, na.rm = TRUE)
max(bill_lengths, na.rm = TRUE)
```
# line 21: Assigning the variable to the vector made from the "bill_length_mm" variable
# lines 23-27: Finding all necessary metrics. "na.rm" makes the calculation ignore missing values in the calculation.

```{r}
# 1.2: Logical subsetting

long_bills <- bill_lengths[which(bill_lengths > 45)] 
long_bills
```
# Used the which function here because "na.rm" does not work in square brackets and I wanted to drop all missing values from the output.

```{r}
# 2.1: Island Analysis
summary(penguins)

dream_bills <- penguins$bill_length_mm[penguins$island == "Dream"] 
mean(dream_bills, na.rm = TRUE) 
```
# line 44: extracts the penguins from Dream island and their bill lengths and saves them to a variable
# line 45: calculates and returns the mean Dream island penguin bill length. ensured to get rid of missing values

```{r}
# 2.2: Species Counting
sum(penguins$species == "Adelie") 
```
# Initially tried to simply assign all adelie penguins to a variable using the same framework as 2.1 above but it returned the word "Adelie" 152 times for all 152 Adelie penguins. Then, used the sum function to return a scalar.

```{r}
# 2.3: Find the heaviest penguin
summary(penguins)
heaviest <- which.max(penguins$body_mass_g) 
penguins$body_mass_g[heaviest]
penguins$species[heaviest]
penguins$island[heaviest]
```


```{r}
# 3.1: Create measurement matrix
penguins_complete <- penguins[complete.cases(penguins), ] # provided

penguin_measurements <- as.matrix(penguins_complete[, c("bill_length_mm", "bill_depth_mm", "flipper_length_mm", "body_mass_g")])
penguin_measurements
```
# Originally tried creating the matrix using just "matrix(penguins_complete) ..." but this doesn't work because it treats each column as one big element. Therefore, I needed to use as.matrix because this is what's able to handle the dataframe properly and return all the elements of each column instead of simply the column name. 

# I also left the space before the comma in the penguin_measurements line blank because each matrix is created in terms of [rows, columns]. I wanted each column to represent the elements in a variable so I left the row area blank and this allows each row to have the element's index.

```{r}
# 3.2: Apply functions
apply(penguin_measurements, 2, mean) # Returns the mean of each column

apply(penguin_measurements, 2, which.max)
# which.max function returns the index of the maximum value per column.
```


```{r}
# 4.1 Write standardize function
standardize <- function(x) {
    x <- x[!is.na(x)] 
    (x - mean(x)) / sd(x)
}

bill_length_standardized <- standardize(penguins$bill_length_mm)
mean(bill_length_standardized)
sd(bill_length_standardized)
```
# explaining function creation: This is a logical vector and will return only the values that are TRUE. In this case, the ! makes the non-NA values TRUE (flips the definition) so it will only return the non-NA values. I saved the vector to a variable so that I can perform operations on the whole vector to return (x - mean) / sd.

```{r}
# Bonus: Option A
adelie <- penguins[penguins$species == "Adelie", ]
chinstrap <- penguins[penguins$species == "Chinstrap", ]
gentoo <- penguins[penguins$species == "Gentoo", ]
# created variables storing the different penguin species.

species_summary <- data.frame(
  species = c("Adelie", "Chinstrap", "Gentoo"), 
  # concatenated the different species in a column
  count = c(nrow(adelie), nrow(chinstrap), nrow(gentoo)), 
  # using nrow returns the number of elements in a row
  mean_body_mass_g = c(mean(adelie$body_mass_g, na.rm = TRUE), 
  # calculated the mean of each different species of penguin separately
  # remembering to ignore missing values using "na.rm"
                       mean(chinstrap$body_mass_g, na.rm = TRUE),
                       mean(gentoo$body_mass_g, na.rm = TRUE)),
  mean_flipper_length_mm = c(mean(adelie$flipper_length_mm, na.rm = TRUE), 
  # calculated the mean of each species' flipper length
  # also ignored missing values using "na.rm"
                             mean(chinstrap$flipper_length_mm, na.rm = TRUE),
                             mean(gentoo$flipper_length_mm, na.rm = TRUE))
)
species_summary
```